---
title: "Data Exploration"
author: "Greg Chism"
format: 
  html:
    theme: flatly
toc: true
editor: visual
---

## Time Series Analysis I: Data Exploration

This document provides R code for the Data Exploration section of lecture 11 from INFO 523 in Fall 2023. The python version of the code can be found

## Install packages

```{r}
if (!require(pacman)) install.packages('pacman')

p_load(colorblindr,
       dlookr,
       here,
       janitor,
       tidyverse,
       timetk)

theme_set(theme_minimal(base_size = 16))
```

## GDP Data

The first timeseries will consider is the US Quarterly GDP for the past 70 years.

```{r}
GDP <- read_csv(here("data", "GDP.csv")) |>
  mutate(DATE = as_date(DATE))
```

### Diagnose the dataset

```{r}
GDP |> diagnose()
```

### Glimpse the dataset

```{r}
GDP |> glimpse()
```

### Plot the GDP as a time series

There is a clear upward trend:

```{r}
GDP |>
  plot_time_series(.date_var = DATE,
                   .value = GDP,
                   .interactive = TRUE,
                   .x_lab = "Date",
                   .y_lab = "GDP ($B)",
                   .smooth = FALSE,
                   .line_color = "#378DBD")
```

## Influenza mortality

Next, we look the percentage of deaths attributed to Influenza and Pneumonia in the past 8 years:

```{r}
ILI <- read_csv(here("data", "CDC.csv")) |>
  clean_names() |>
  mutate(date = year + week/52)

```

### Glimpse the data

```{r}
ILI |> glimpse()
```

### Plot Influenza mortality as a time series

The seasonal behavior in unmistakable. The purple curve illustrates the expected seasonal behavior, while the green curve displays the threshold above which we are facing an epidemic:

```{r}
ILI |>
  select(date, percent_of_deaths_due_to_pneumonia_and_influenza, expected, threshold) |>
  pivot_longer(!date, names_to = "series", values_to = "values") |>
  mutate(series = fct_relevel(series, "percent_of_deaths_due_to_pneumonia_and_influenza",
                              "expected", "threshold")) |>
  ggplot(aes(x = date, y = values, color = series)) +
  geom_line(size = 1) + 
  scale_color_manual(labels = c("% Mortality", "Expected", "Threshold"),
                       values = c("#378DBD", "#E69F00", "#009E73")) +
  labs(x = "Date", y = "% Mortality") +
  theme(legend.position = "top")
  

```

## Sun activity

The number of observed sun spots over 300 years also display a nice seasonal behavior over the course of a decade or so:

```{r}
Sun <- read_csv(here("data", "sun.csv")) 
```

### Glimpse the data

```{r}
Sun |> glimpse()
```

```{r}
Sun |>
ggplot(aes(x = YEAR, y = SUNACTIVITY)) +
  geom_line(size = 1, color = "#378DBD") +
  labs(x = "Year", y = "Sun spot activity") 
```

## Dow-Jones Industrial Average

The Dow-Jones Industrial Average also has a clear uptrend like GDP but with significantly larger fluctuations:

```{r}
DJIA <- read.csv(here("data", "DJIA.csv"), na.strings = ".") |>
  mutate(DATE = as_date(DATE))
```

### Glimpse the data

```{r}
DJIA |> glimpse()
```

### Diagnose the data

```{r}
DJIA |> diagnose()
```

```{r, warning=FALSE}
DJIA |>
  filter(between(DATE, as.Date("2016-12-31"), as.Date("2019-01-01"))) |>
  ggplot(aes(x = DATE, y = DJIA)) +
  geom_line(size = 1, color = "#378DBD") +
  labs(x = "Date", y = "DIJA") +
  scale_x_date(
    date_breaks = "4 months",
    labels = c("Jan\n2017", "Apr", "Jul", "Oct", "Jan\n2018", "Apr", "Jul", "Oct")
  ) +
  coord_cartesian(xlim = as.Date(c("2016-12-31", "2019-01-01"))) 
```

## Airline Passengers

The number of passenters in international flights in the 50s displays both a trend and a seasonal pattern

```{r}
airline <- read.csv(here("data", "international-airline-passengers.csv"), sep = ";") |>
  mutate(Month = paste0(Month, "-01"),
         Month = as_date(Month))
```

```{r}
airline |> head()
```

```{r}
airline |>
  ggplot(aes(x = Month, y = Passengers)) +
  geom_line(size = 1, color = "#378DBD") +
  labs(x = "Date", y = "DIJA") +
  scale_x_date(date_breaks = "2 years", 
               date_labels = "%Y")
```

## Types of non-stationarity

There are several kinds of non-stationarity. Here we illustrate the most common ones with simple toy examples:

```{r}
# Create data
x <- seq(0, 10*pi, length.out = 360)
y <- sin(x)
df <- tibble(
  x = x,
  stationary = y,
  changing_mean = y + x/10,
  changing_variance = y * x/10,
  changing_covariance = sin(x + x^2/30)
)

# Plot
plots <- list(
  ggplot(df, aes(x = x, y = stationary)) +
    geom_line(size = 1, color = "#378DBD") +
    labs(title = "Stationary series", x = "time", y = "Amplitude"),
  
  ggplot(df, aes(x = x, y = changing_mean)) +
    geom_line(size = 1, color = "#378DBD") +
    labs(title = "Changing mean", x = "time", y = "Amplitude"),
  
  ggplot(df, aes(x = x, y = changing_variance)) +
    geom_line(size = 1, color = "#378DBD") +
    labs(title = "Changing Variance", x = "time", y = "Amplitude"),
  
  ggplot(df, aes(x = x, y = changing_covariance)) +
    geom_line(size = 1, color = "#378DBD") +
    labs(title = "Changing Co-variance", x = "time", y = "Amplitude")
)

# Arrange the plots in a grid
gridExtra::grid.arrange(plots[[1]], plots[[2]], plots[[3]], plots[[4]], ncol = 2)

```

## Kinds of trend

Two of the most common non-stationarity types are due to differnet kinds of trends that we illustrate here:

```{r}
# Create data
x <- seq(0, 10*pi, length.out = 360)
y <- sin(x)
df <- tibble(
  x = x,
  additive_trend = y + x/10,
  multiplicative_trend = y * x/10
)

# Additive Trend Plot
additive_plot <- ggplot(df, aes(x = x, y = additive_trend)) +
  geom_line(size = 1, color = "#378DBD") +
  labs(title = "Additive Trend", x = "time", y = "Amplitude") 

# Multiplicative Trend Plot
multiplicative_plot <- ggplot(df, aes(x = x, y = multiplicative_trend)) +
  geom_line(size = 1, color = "#378DBD") +
  labs(title = "Multiplicative Trend", x = "time", y = "Amplitude") 

# Arrange the plots side by side
gridExtra::grid.arrange(additive_plot, multiplicative_plot, ncol = 2)
```
